---
title: "Fitting Exercise"
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: console
---

# Introduction

For this assignment, I will be using data on a drug candidate called
Mavoglurant from the paper:

Wendling, T., Dumitras, S., Ogungbenro, K. et al. Application of a
Bayesian approach to physiological modelling of mavoglurant population
pharmacokinetics. J Pharmacokinet Pharmacodyn 42, 639–657 (2015).
https://doi.org/10.1007/s10928-015-9430-4

## Loading

First we will load the data and additional packages necessary for the
assigniment.

```{r}
# Installing package of data
library(nlmixr2data)

# Installing additional packages
library(here)
library(tidyr)
library(ggplot2)
library(readxl)
```

I will also load the data from the another paper which has used this
data to ensure a level of consistency and ease of usage with the data.

```{r}
# Loading Data through CSV
mavodrug <-- read.csv('Mavoglurant_A2121_nmpk.csv')
```

# Exploring and Processing

```{r}
# Change all values in the entire data frame to positive
mavodrug <- abs(mavodrug)

# Check column names
colnames(mavodrug)

# Check data types
str(mavodrug)
```

This is time-series data of drug concentrations. So for further
exploration, we will consider time as the independent variable. To
visualilze the data, We will plot the outcome variable (DV) as a
function of time, stratified by DOSE and using ID as a grouping factor.

```{r}
# Plotting Outcome Variable

# Convert DOSE and ID to factors if they are not already
mavodrug$DOSE <- factor(mavodrug$DOSE)
mavodrug$ID <- factor(mavodrug$ID)

# Plotting
ggplot(mavodrug, aes(x = TIME, y = DV, color = DOSE, group = ID)) +
  geom_line() +
  labs(title = "DV over time stratified by dose and ID",
       x = "Time",
       y = "DV") +
  scale_color_manual(values = c("red", "green", "blue")) +
  theme_minimal()
```

There are some individuals that have received the drug more than once,
indicated by having both entries with OCC=1 and OCC=2. To keep things
simple, I only keep one dataset for each individual and remove all
entries with OCC=2.

```{r}
# Checking unique values in the 'OCC' column
unique(mavodrug$OCC)

# Filter observations with OCC = 1
mavodrug <- mavodrug[mavodrug$OCC == 1, ]

# Check the structure of the data frame after filtering
str(mavodrug)
```

Here, I will compute the total amount of drug for each individual (for
exercise purposes only, not used in normal practices).

```{r}
library(dplyr)

# Exclude observations with TIME = 0
mavodrug_filtered <- mavodrug %>%
  filter(TIME != 0)

# Compute the sum of the DV variable for each individual
summarized_mavo <- mavodrug_filtered %>%
  group_by(ID) %>%
  summarize(Y = sum(DV))

# Create a data frame containing only the observations where TIME == 0
time_0_mavo <- mavodrug %>%
  filter(TIME == 0)

# Use the appropriate join function to combine the two data frames
combined_mavo <- inner_join(summarized_mavo, time_0_mavo, by = "ID")

# Check the dimensions of the resulting data frame
dim(time_0_mavo)
dim(combined_mavo)
```

```{r}
# Convert RACE and SEX to factor variables
combined_mavo <- combined_mavo %>%
  mutate(RACE = factor(RACE),
         SEX = factor(SEX))

# Select only the specified variables and check the structure
cleaned_mavo <- combined_mavo %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

# Check the structure of the cleaned data
str(cleaned_mavo)
```

## Summary Tables

The summary tables provide a quick overview of the numerical and
categorical variables in the dataset.

```{r}
# Summary table for numerical variables
summary_table_numeric <- summary(cleaned_mavo[, c("Y", "AGE", "WT", "HT")])

# Summary table for categorical variables
summary_table_categorical <- table(cleaned_mavo$DOSE, cleaned_mavo$SEX)

# Print summary tables
print("Summary table for numerical variables:")
print(summary_table_numeric)

print("Summary table for categorical variables:")
print(summary_table_categorical)
```

## Plots

These plots show the relationship between the total drug ('Y') and other
predictors such as age, dose, and sex. Scatterplots help visualize the
continuous relationship between 'Y' and age, while boxplots illustrate
the distribution of 'Y' across different levels of dose and sex.

```{r}
# Scatterplot of Y vs. AGE
plot(Y ~ AGE, data = cleaned_mavo, main = "Total Drug vs. Age", xlab = "Age", ylab = "Total Drug")

# Boxplot of Y by DOSE
boxplot(Y ~ DOSE, data = cleaned_mavo, main = "Total Drug by Dose", xlab = "Dose", ylab = "Total Drug")

# Boxplot of Y by SEX
boxplot(Y ~ SEX, data = cleaned_mavo, main = "Total Drug by Sex", xlab = "Sex", ylab = "Total Drug")
```

## Distributions of Variables

These histograms visualize the distributions of variables such as total
drug ('Y'), age, weight ('WT'), and height ('HT'). They help identify
any potential outliers or unusual patterns in the data.

```{r}
# Histogram of Y
hist(cleaned_mavo$Y, main = "Distribution of Total Drug", xlab = "Total Drug")

# Histogram of AGE
hist(cleaned_mavo$AGE, main = "Distribution of Age", xlab = "Age")

# Histogram of WT
hist(cleaned_mavo$WT, main = "Distribution of Weight", xlab = "Weight")

# Histogram of HT
hist(cleaned_mavo$HT, main = "Distribution of Height", xlab = "Height")
```

## Pair/Correlation Plots

The pair plot provides a visual overview of the relationships between
the variables 'Y', age, weight, and height. The correlation matrix
quantifies the strength and direction of the linear relationships
between these variables. It helps identify potential multicollinearity
issues and informs feature selection for modeling purposes.

```{r}
# Pair plot
pairs(cleaned_mavo[, c("Y", "AGE", "WT", "HT")])

# Correlation matrix
correlation_matrix <- cor(cleaned_mavo[, c("Y", "AGE", "WT", "HT")])
print("Correlation matrix:")
print(correlation_matrix)
```

# Model Fitting

Here, I conduct a model fitting to assess the relationship of the
outcome of interest with the other variables. Tidymodels provides the
functions necessary for modeling and preprocessing data. The recipe
function specifies the data preprocessing steps. In this case, the
outcome variable (Y) is defined as the response variable, and all other
variables are considered predictors. step_dummy converts categorical
predictors into dummy variables, step_center centers numeric predictors
around their mean, and step_scale scales numeric predictors to have unit
variance.

```{r}
# Load necessary libraries
library(tidymodels)
# Helper packages
library(readr)       # for importing data
library(broom.mixed) # for converting bayesian models to tidy tibbles
library(dotwhisker)  # for visualizing regression results
library(tidymodels)

# Define the recipe
mavo_recipe <- recipe(Y ~ ., data = cleaned_mavo) %>%
  step_dummy(all_nominal(), -all_outcomes()) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())
```

The initial_split function divides the dataset into training and testing
sets. The prop argument specifies the proportion of the data to allocate
to the training set (80% in this case), and the strata argument ensures
that the split is stratified based on the outcome variable (Y), which
helps maintain the balance of categories in both sets.

The linear_reg function specifies the linear regression model.
set_engine("lm") selects the linear model engine (ordinary least squares
regression), and set_mode("regression") sets the mode of the model to
regression, indicating that it predicts a continuous outcome (Y).

The workflow function creates a modeling workflow. This workflow
combines the preprocessing steps defined in the recipe with the
specified model. The add_recipe function adds the recipe to the
workflow, and the add_model function adds the linear regression model.

```{r}
# Split the data into training and testing sets
set.seed(123) # for reproducibility
mavo_split <- initial_split(cleaned_mavo, prop = 0.8, strata = Y)
mavo_train <- training(mavo_split)
mavo_test <- testing(mavo_split)

# Define the linear regression model specification
linear_spec <- linear_reg() %>%
  set_engine("lm") %>%
  set_mode("regression")

# Create a workflow that incorporates the recipe and the model
mavo_workflow <- workflow() %>%
  add_recipe(mavo_recipe) %>%
  add_model(linear_spec)
```

The fit function fits the workflow to the training data. This step
applies the preprocessing steps defined in the recipe to the training
data and then fits the linear regression model to the preprocessed data.

The predict function generates predictions from the fitted model using
the testing data. This step applies the preprocessing steps (from the
recipe) to the testing data and then predicts the outcome variable (Y)
using the fitted linear regression model.

The metrics function computes evaluation metrics for the model
predictions. Here, we calculate the root mean squared error (RMSE) and
the R-squared value to assess the model's performance. The computed
metrics are then extracted and stored in the variables all_rmse and
all_r_squared, respectively.

```{r}
# Fit the workflow to the training data
mavo_fit <- fit(mavo_workflow, data = mavo_train)

# Compute predictions on the testing data
mavo_predictions <- predict(mavo_fit, new_data = mavo_test) %>%
  bind_cols(mavo_test)
```

The metrics function computes evaluation metrics for the model
predictions. Here, we calculate the root mean squared error (RMSE) and
the R-squared value to assess the model's performance. The computed
metrics are then extracted and stored in the variables all_rmse and
all_r_squared, respectively.

```{r}
# Compute RMSE and R-squared for the model using all predictors
all_metrics <- mavo_predictions %>%
  metrics(truth = Y, estimate = .pred)

all_rmse <- all_metrics %>%
  filter(.metric == "rmse") %>%
  pull(.estimate)

all_r_squared <- all_metrics %>%
  filter(.metric == "rsq") %>%
  pull(.estimate)
```

Finally, we print the computed RMSE and R-squared values to evaluate the
model's performance. These metrics provide insights into how well the
linear regression model fits the data and predicts the outcome variable.

```{r}
# Print RMSE and R-squared for the model using all predictors
cat("Model using all predictors:\n")
cat("RMSE:", all_rmse, "\n")
cat("R-squared:", all_r_squared, "\n")
print(mavo_predictions)
print(mavo_fit)
```

# Results

The RMSE is approximately 672.59. Lower values of RMSE indicate better
model performance, as they imply smaller errors between predicted and
actual values. The R-squared value is approximately 0.47, suggesting
that the model explains about 47% of the variance in the outcome
variable. The intercept term is 2459.86, representing the predicted
value of Y when all predictors are set to zero. The coefficient for AGE
is 52.83, indicating that for each one-unit increase in age, the
predicted value of Y increases by 52.83 units. Finally, coefficients for
other predictors (WT, HT, DOSE, SEX, RACE) represent the change in the
predicted value of Y associated with a one-unit change in each
respective predictor.
