---
title: "Data Analysis Exercise"
editor: 
  markdown: 
    wrap: 72
---


When I first downloaded my data in, the size of the file was 33 MB large. R was unable to load the file so I decided to import the data. After importing the data, I wrote the code beloww to reduce the size of the to 1000 observations. I created the reduced set name (reduced_data) and cleaned the data with the following code. 

#This chunk only works in the R terminal as I cannot figure out how it will work within the qmd file. 
```{r}
library(readr)
library(dplyr)

original_data <- read_csv("Nutrition__Physical_Activity__and_Obesity_-_Behavioral_Risk_Factor_Surveillance_System_20240207.csv")

# Define the desired number of rows to include in the reduced dataset
desired_rows <- 1000  # Adjust this number as needed

# Select a random subset of rows
reduced_data <- original_data %>% 
  sample_n(desired_rows)

# Write the reduced dataset to a new CSV file
write_csv(reduced_data, "Nutrition__Physical_Activity__and_Obesity_-_Behavioral_Risk_Factor_Surveillance_System_20240207.csv")

# Retrieve the full column specification
column_spec <- spec(reduced_data)
column_spec
```

I, then, created code to clean the data removing misssing values and changing them to NA. 

```{r}
library(readr)
library(dplyr)

# Display the structure of the data
str(reduced_data)

# Check for missing values
missing_values <- sum(is.na(reduced_data))
cat("Number of missing values in the dataset:", missing_values, "\n")

# Replace missing values (if coded as 999) with NA
reduced_data <- reduced_data %>%
  mutate_all(~ ifelse(. == 999, NA, .))
```
In order to view only a few variables in the data out of the 33 available, I decided that Class, Question, the data value attached to the question, education income and race/ethnicity were pertinent variables. Geographical location was kept as a separate viariable to analyze. 

```{r}
# Select a few variables for analysis (e.g., first 5 variables)
selected_variables <- reduced_data %>%
  select(LocationDesc, Class, Question, Data_Value, Education, Income, `Race/Ethnicity`)

# Display summary statistics for selected variables
summary(selected_variables)
```
To explore the data further, I ran code to assess the distribution of categorical variables below. 

```{r}
library(dplyr)
library(ggplot2)
# Perform exploratory analysis
# Summary table for categorical variables with percentages
categorical_summary <- reduced_data %>%
  summarize(
    LocationDesc_percent = n_distinct(LocationDesc) / n() * 100,
    Class_percent = n_distinct(Class) / n() * 100,
    Question_percent = n_distinct(Question) / n() * 100,
    Education_percent = n_distinct(Education) / n() * 100,
    Income_percent = n_distinct(Income) / n() * 100,
    Race_Ethnicity_percent = n_distinct(`Race/Ethnicity`) / n() * 100
  )

# Summary statistics for continuous variable
continuous_summary <- reduced_data %>%
  summarize(
    Data_Value_mean = mean(Data_Value, na.rm = TRUE),
    Data_Value_sd = sd(Data_Value, na.rm = TRUE)
  )

# Print summary tables
print("Summary Table for Categorical Variables:")
print(categorical_summary)
cat("\n")

print("Summary Table for Continuous Variables:")
print(continuous_summary)
```

Here, I created a plot to show the ditribution of data by physical activity class and filtered out the question "Percent engaging in no leisure-time physical activity" to look at the distibution of data within this class and question. 

```{r}
library(dplyr)

physical_activity_data <- reduced_data %>%
  filter(Class == "Physical Activity")

physical_activity_no_leisure_data <- reduced_data %>%
  filter(Class == "Physical Activity" & 
         Question == "Percent of adults who engage in no leisure-time physical activity")

library(ggplot2)

# Summary table for categorical variables
categorical_summary <- physical_activity_no_leisure_data %>%
  summarise(across(c(LocationDesc, Education, Income, `Race/Ethnicity`), ~round(mean(.)*100, 2))) %>%
  rename(LocationDesc_Percentage = LocationDesc,
         Education_Percentage = Education,
         Income_Percentage = Income,
         Race_Ethnicity_Percentage = `Race/Ethnicity`)

print("Summary of Categorical Variables:")
print(categorical_summary)

# Histogram and summary statistics for the continuous variable Data_Value
ggplot(physical_activity_no_leisure_data, aes(x = Data_Value)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Percentages of Adults\nEngaging in No Leisure-Time Physical Activity",
       x = "Percentage of Adults",
       y = "Frequency") +
  theme_minimal()

mean_data_value <- mean(physical_activity_no_leisure_data$Data_Value)
sd_data_value <- sd(physical_activity_no_leisure_data$Data_Value)

print("Summary of Continuous Variable (Data_Value):")
print(paste("Mean:", round(mean_data_value, 2)))
print(paste("Standard Deviation:", round(sd_data_value, 2)))
```

